---
title: "'MaxentVariableSelection' vignette" 
author: "Alexander Jueterbock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{'MaxentVariableSelection' vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

[//]: (Put here a table of contents)


[//]: Inform how to cite this package [//]: Refer in the details
[//]: section of the VariableSelection.Rd file to this vignette and don't
[//]: mention there all details.

[//]: Refer in the examples section of the package to this
[//]: vignette. Also, remove the sstmax variable from the package because it takes too much space



# Introduction

Complex niche models show low performance in identifying the most
important range-limiting environmental variables and in transferring
habitat suitability to novel environmental conditions (Warren and
Seifert, 2011; Warren *et al*., 2014). This vignette demonstrates how to
constrain complexity and increase perforance of Maxent niche models by
identifying the most important set of uncorrelated environmental
variables and by fine-tuning Maxent’s regularization multiplier.


# Requirements and input data

Users of this package should be familiar with Maxent Niche
Modelling. If you are not, you can find a great tutorial
[here](https://www.google.no/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0CB4QFjAAahUKEwjPwrHpxbjHAhVHliwKHXc6Ai8&url=https%3A%2F%2Fwww.cs.princeton.edu%2F~schapire%2Fmaxent%2Ftutorial%2Ftutorial.doc&ei=oT7WVY_uIMessgH39Ij4Ag&usg=AFQjCNGLRzLmkwCWYL3VM30dFpe3QBLSbQ&sig2=alOYlorHojDHaHA-v3PnmA).


## ASCII Grids of environmental variables

To test for the importance of environmental variables in
discriminating suitable from non-suitable habitat, these variables
must be available as ASCII grids
([in ESRI’s .asc format](https://en.wikipedia.org/wiki/Esri_grid)). All
variables must have the same extent and resolution.

For example, biologically meaningful variables are freely available
from the [WorldClim dataset](http://www.worldclim.org/bioclim)
(Hijmans *et al*., 2005) for terrestrial systems and from the
[Bio-ORACLE dataset](http://www.oracle.ugent.be/) (Tyberghein *et
al*. 2012) for marine systems. 


## Maxent jar file

This package runs Maxent to evaluate the performance of models built
with different subsets of environmental variables. The Maxent program,
however, doesn't come with this R package. It has to be downloaded
separately from
[here](https://www.cs.princeton.edu/~schapire/maxent/). The program is
free but you will have to enter your name, institution, and email
prior to downloading. This package has been tested with Maxent
version 3.3.3k.


## Files of occurrence and background locations

Geocoded background and occurrence records have to be provided in a
format that is referred to in the
[Maxent tutorial](https://www.google.no/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0CB4QFjAAahUKEwjPwrHpxbjHAhVHliwKHXc6Ai8&url=https%3A%2F%2Fwww.cs.princeton.edu%2F~schapire%2Fmaxent%2Ftutorial%2Ftutorial.doc&ei=oT7WVY_uIMessgH39Ij4Ag&usg=AFQjCNGLRzLmkwCWYL3VM30dFpe3QBLSbQ&sig2=alOYlorHojDHaHA-v3PnmA)
as SWD format. The datasets `Backgrounddata.csv` and
`Occurrencedata.csv`, which were used in Jueterbock2015 and are
included in this package, exemplify the required content.

For each location, this file lists the longitude and latitude values,
as well as the values for each environmental variable.
[//]:The values are
[//]:separated by commas, as in these example lines, with the first line as header line:

[//]:```
[//]:species,longitude,latitude,calcite,salinity,maxtemperature
[//]:bg,39.79,65.29,0.008,34.577,10.28 bg,21.12,79.54,0.014,22.214,1.28
[//]:bg,-4.95,55.29,0.001,29.077,15.64 bg,-135.37,59.29,0.006,31.59,12.05
[//]:bg,18.04,42.62,0.001,33.575,25.93 bg,-4.45,48.70,0.003,31.614,15.62
[//]:```

Let's have a look at the content of `Occurrencedata.csv`:

```{r}
occurrencelocations <- system.file("extdata", "Occurrencedata.csv", package="MaxentVariableSelection")
occurrencelocations <- read.csv(occurrencelocations,header=TRUE)
head(occurrencelocations)
```

[//]: Does this show the output?

For each of 98 occurrence records of the macroalga *Fucus distichus*
(`Fd`), this dataset contains `longitude` and `latitude` values, as
well as values for four environmental variables (`calcite`, `parmean`,
`salinity`, and `sstmax`). 


Such files are easy to create with the *extract* function of the R
package 'raster' (Hijmans, 2015) once we have longitude and latitude
values of the occurrence records.

You can install the 'raster' package with the following command

```{r, eval=FALSE}
install.packages('raster')
```

Let's extract variables from the
Bio-ORACLE database (Tyberghein et al., 2012) for each of the 98
occurrence records of *Fucus distichus*. Downloading the variables
(from [here](www.oracle.ugent.be/DATA/90_90_RV/BioOracle_9090RV.rar))
may take a while as the rasters sum to about 160 MB in size.  the
variables are compressed as RAR files. Extract them and save them in a
folder called `BioORACLEVariables`. The following R code shows how to
obtain the value of these variables for each of the occurrence records

```{r} 
# load the raster package
library(raster)

# Load the occurrence records 
occurrencelocations <- system.file("extdata", "Occurrencedata.csv", package="MaxentVariableSelection")
occurrencelocations <- read.csv(occurrencelocations,header=TRUE)
LonLatData <- occurrencelocations[,c(2,3)]

# Then load the environmental variables into R with the help of the
# stack function of the 'raster' package. You can not just copy the
# following line but have to adjust the filepath to your own. 
files <- list.files("/home/alj/Downloads/BioORACLEVariables",pattern='asc',
full.names=TRUE)
Grids <- raster::stack(files)

# Extracting the variables for all occurrencelocations
VariablesAtOccurrencelocations <- raster::extract(Grids,LonLatData)

# Combining the extracted values with the longitude and latitude values
Outfile <- as.data.frame(cbind("Fucusdistichus",LonLatData,VariablesAtOccurrencelocations))
colnames(Outfile) <- c("species","longitude","latitude",colnames(VariablesAtOccurrencelocations))


#writing this table to a csv file:

write.csv(Outfile, file =
"VariablesAtOccurrencelocations.csv", append = FALSE,sep = ",", eol =
"\n", na = "NA", dec = ".",col.names = TRUE,row.names=FALSE)
```

# Workflow tutorial

The central function of this package is called
`VariableSelection`. The basic usage of this function is:

```{r, eval=FALSE}
VariableSelection(maxent, outdir, gridfolder,
occurrencelocations, backgroundlocations, additionalargs,
contributionthreshold, correlationthreshold, betamultiplier)
```


The following sections guides you through an example with the possible
settings of each argument.





## `maxent`

Specify the file path to the `maxent.jar` file, which has to be
downloaded from
[here](https://www.cs.princeton.edu/~schapire/maxent/). Not that you
can not just copy the following line but instead have to specify the
file path to the folder containing `maxent.jar` on your own computer.

```{r}
maxent <- ("/home/alj/Downloads/maxent.jar")

```

If you are working on a Windows platform, your filepath might rather
look similar to the following, where ... should be replaced by the
hierarchical file structure that leads to `maxent.jar` on your
computer.:

```{r, eval=FALSE}
maxent <- ("C:/.../maxent.jar")
```


## `outdir`
Filepath to the output directory, including the name of the directory, like:

```{r}
outdir <- ("/home/alj/Downloads/OutputDirectory")
```

Again, you need to adjust the filepath to the correct path and folder
on your own computer. All result files will be written into this
folder. Please don't put important files in this folder as all files
but the output files of the `VariableSelection` function will be
deleted from this folder.



## `gridfolder`
	
Here, you specify the filepath to the folder containing all your ASCII
grids of environmental variables that you consider to be potentially
relevant in setting distribution limits of your target species.
All variables must have the same extent and resolution.

If you are puzzled what grids I am referring to, have a look at the
section **ASCII Grids of environmental variables** above.

Like for the `maxent` and `outdir` arguments, you also can not just
copy the following line as the location of ASCII grids is likely
different on your computer. Adjust the filepath to the folder where
you stored your ASCII grids. In this example I provide the filepath to
the `Bio-ORACLE` rasters that I downlaoded from
[here](www.oracle.ugent.be/DATA/90_90_RV/BioOracle_9090RV.rar).

```{r}
gridfolder <- ("/home/alj/Downloads/BioORACLEVariables")
```

## `occurrencelocations`

Here, you need to specify the filepath to the csv file of occurrence
locations (see the section **Files of occurrence and background
locations** above for the required file format). An example file of
occurrence locations for the macroalga *Fucus distichus* is included
in this package. You load it with:

```{r}
occurrencelocations <- system.file("extdata", "Occurrencedata.csv", package="MaxentVariableSelection")
```

Instead, if the file with your occurrece locations is stored on your
own computer, set here the filepath to it. For example:

```{r, eval=FALSE}
occurrencelocations <- "/home/alj/Downloads/Occurrencedata.csv"
```



## `backgroundlocations`

The same applies for the `backgroundlocations` argument as for the
`occurrencelocations` argument.

If you want to use the background locations that were used in (XX specify the submitted paper), type:

```{r}
backgroundlocations <- system.file("extdata", "Backgrounddata.csv", package="MaxentVariableSelection")
```

Instead, if you have your own csv file with backgroundlocations
specify the filepath here, similar to:

```{r, eval=FALSE}
backgroundlocations <- "/home/alj/Downloads/Backgrounddata.csv"
```

## `additionalargs`

Maxent arguments can be specified with the `additionalargs` argument.
You find an overview of possible arguments in the end of the Maxent
help file.
  
The following settings are preset in this package to calculate
information criteria and AUC values. Don't change these values in the
`additionalargs` string:

- `autorun=true`: starts running with the startup of the program.
- `plots=false`: does not create any plots.
- `writeplotdata=false`: does not write output data that are required
  to plot response curves.
- `visible=false`: does not show the Maxent user interface but runs the
  programm from the command line.
- `randomseed=true`: the background data will be split into different
  test/train partitions with each new run.
- `writebackgroundpredictions=false`: does not write .csv files with
  predictions at background points.
- `replicates=10`: number of replicate runs; this is only set for the
  calculation of AUC values. The calculation of AICc values doesn't
  require to partition the occurrence sites in test and training data,
  thus, also does not require to replicate runs.
- `replicatetype=subsample`: uses random test/training data for
  replicated sample sets; this is only set for the calculation of AUC
  values.
- `randomtestpoints=50`: 50 percent of presence points are randomly
  set aside as test data; this is only set for the calculation of AUC
  values.
- `redoifexists`: redoes the analysis if output files exist already.
- `writemess=false`: does not write multidimensional similarity
  surfaces (MESS), which inform on novel climate conditions in the
  projection layers.
- `writeclampgrid=false`: does not write a grid on the spatial
  distribution of clamping.
- `askoverwrite=false`: does not ask if the data shall be
  overwritten if they exist already.
- `pictures=false`: does not create .png images of the output grids.
- `outputgrids=false`: does not write output grids for all replicate
  runs but only the summary grids; this is only set for the
  calculation of AUC values.
- `outputformat=raw`: write output grids with raw probabilities of
  presence; this is only set for calculation of information criteria,
  like AICc.


Instead, what additional Maxent argument you might want to set are the
features to be used, by selecting true or false for the arguments
`linear`, `quadratic`, `product`, `threshold`, and `hinge`.

For this example, we are using only `hinge` features, with the
following setting:

```{r}
additionalargs="nolinear noquadratic noproduct nothreshold noautofeature"
``` 

## `contributionthreshold`

This sets the threshold of model contribution below which
environmental variables are excluded from the Maxent model. Model
contributions range from 0% to 100% and reflect the importance of
environmental variables in limiting the distribution of the target
species.

In this example, we set the threshold to 5%, which means that all
variables will be excluded when they contribute less than 5% to the
model:

```{r}
contributionthreshold <- 5 
```

## `correlationsthreshold`

This sets the threshold of Pearson's correlation coefficient (ranging
from 0 to 1) above which environmental variables are regarded to be
correlated (based on values at all background locations). Of the
correlated variables, only the variable with the highest contribution
score will be kept, all other correlated variables will be excluded
from the Maxent model. Correlated variables should be removed because
they may reflect the same environmental conditions, and can lead to
overly complex or overpredicted models. Also, models compiled with
correlated variables might give wrong predictions in scenarios where
the correlations between the variables differ.

Here, we are setting the threshold to 0.9:

```{r]
correlationthreshold <- 0.9
```


## `betamultiplier`

This argument sets the values of beta multipliers (regularization
multipliers) for which variable selection shall be performed. The
smaller this value, the more closely will the projected distribution
fit to the training data set. Models that are overfitted to the
occurrence data are poorly transferable to novel environments and,
thus, not appropriate to project distribution changes under
environmental change. Performance will be compared between models
created with the beta values given in this `betamultiplier`
vector. Thus, providing a range of beta values from 1 (the default in
Maxent) to 15 or so, will help you to spot the optimal beta multiplier
for your specific model.

Here, we are testing betamultipliers from 4 to 6 with steps of 0.5
with the following setting: 

```{r}
 betamultiplier=seq(4,6,0.5)
```

## Variable selection procedure

With the argument seetings that we specified above, we can start the
variable selection.  The following function starts the model-selection
procedure

```{r, eval=FALSE}
library("MaxentVariableSelection")
VariableSelection(maxent,
                  outdir,
                  gridfolder,
                  occurrencelocations,
                  backgroundlocations,
                  additionalargs,
                  contributionthreshold,
                  correlationthreshold,
                  betamultiplier
                  )
```

The run will take a while (ca. half an hour).  The resulting output
files will be saved in the folder that you specified in the argument
`outdir`.

The following paragraphs explain what steps are actually taken in this
procedure to identify the set of most relevant environmental
variables.  First, an initial Maxent model is compiled with all
provided environmental variables. Then, all variables are excluded
which have a relative contribution score below the value set with
`contributionthreshold`.  Then, those variables are removed that
corrleate with the variable of highest contribution (at a correlation
coefficient > `correlationthreshold` or < -`correlationthreshold`).

The remaining set of variables is then used to compile a new Maxent
model. Again, variables with low contribution scores are removed and
remaining variables that are correlated to the variable of
second-highest contribution are discarded. This process is repeated
until left with a set of uncorrelated variables that all had a model
contribution above the value set with contributionthreshold. These
steps are perfored for the range of betamultipliers specified with
`betamultiplier`.

The performance of each Maxent model is assessed with the
sample-size-adjusted Akaike information criterion (AICc) (XXAkaike,
1974), the area under the receiver operating characteristic (AUC)
estimated from test data (XXFielding and Bell, 1997), and the
difference between AUC values from test and training data (AUC.Diff),
which estimates model overfitting. AICc values are estimated from
single models that include all occurrence sites, based on code in
ENMTools (xxWarren et al., 2010). AUC values, instead, are averaged
over ten replicate runs which differ in the set of 50% test data that
are randomly sub-sampled from the occurrence sites and withheld from
model construction.

Maximization of test AUC values generally favors models that can well
discriminate between presence and absence sites (Fielding and Bell,
1997; Warren and Seifert, 2011). This, however, bears the risk to
overpredict the realized niche of a species (Jimenez-Valverde, 2012).
Instead, minimization of AICc values generally favors models that
recognize the fundamental niche of a species and that are better
transferable to future climate scenarios (Warren and Seifert, 2011).
The optimal set of variables along with the best beta multiplier is
then identified as the model of highest AUC or lowest AICc values (the
results of the `VariableSelection` functions gives you both options).





# Results

The results are saved in the directory that was specified in the
argument `outdir`. If you were patient enough to wait for the example
variable selection to finish (see above), you will find seven tables
and four figures in the output directory.

ModelPerformance.txt
A table listing the performance indicators of all created Maxent models

ModelWithMaxAUCTest.txt
Subset of the table ModelPerformance.txt,

ModelWithMinAICc.txt
Subset of the table ModelPerformance.txt, which shows only the model with
the lowest AICc value.


X_VariableSubset.csv
occurrencelocations and backgroundlocations input files reduced to the
final set of selected environmental variables.
XX  But based on AICc or AUC? Specify this also in the vignette


VariableSelectionProcess.txt
Table listing model contributions for and correlations between each of the envi-
ronmental variables for all created Maxent models.

VariableSelectionMaxAUCTest.txt
Subset of VariableSelectionProcess.txt that shows only those models which
lead directly to the model with the highest AUC.Test value.

VariableSelectionMinAICc.txt
Subset of VariableSelectionProcess.txt that shows only those models which
lead directly to the model with the lowest AICc value.





ModelSelectionAICc_MarkedMaxAUCTest.png
A figure showing the AICc values of all models, which are ordered along the
x-axis based on the applied beta-multiplier.

ModelSelectionAICc_MarkedMinAICc.png
A figure showing the AICc values of all models, which are ordered along the
x-axis based on the applied beta-multiplier.

ModelSelectionAUCTest_MarkedMaxAUCTest.png
A figure showing the AUC.Test values of all models, which are ordered along
the x-axis based on the applied beta-multiplier.

Describe the example results here now

Explain why tohe maximum number of variables is four and why not all
of the given variables are taken for modelling (because they are not
specified in the occurrencelocations and backgroundlocations files

XX 20150829ContinueHere
 show the resulting output files and tables to explain them.

ModelPerformance.txt
# show here also directly the figures


# References
Hijmans, R.J., Cameron, S.E., Parra, J.L., Jones, P.G. & Jarvis, A.
(2005) Very high resolution interpolated climate surfaces for
global land areas. International Journal of Climatology, 25,
1965–1978

Tyberghein, L., Verbruggen, H., Pauly, K., Troupin, C., Mineur, F. and
De Clerck, O. (2012), Bio-ORACLE: a global environmental dataset for
marine species distribution modelling. Global Ecology and
Biogeography, 21: 272–281. doi: 10.1111/j.1466-8238.2011.00656.x

Warren, D. L., & Seifert, S. N. (n.d.). Ecological niche modeling in
Maxent: the importance of model complexity and the performance of
model selection criteria. Ecological Applications, 21(2), 335–342.

Warren, D. L., Wright, A. N., Seifert, S. N., & Shaffer, H. B. (2014). Incorporating model complexity and spatial sampling bias into ecological niche models of climate change risks faced by 90 California vertebrate species of concern. Diversity and Distributions, 20(3), 334–343. doi:10.1111/ddi.12160
[//]: (http://rmarkdown.rstudio.com/)


  @Manual{,
    title = {raster: Geographic Data Analysis and Modeling},
    author = {Robert J. Hijmans},
    year = {2015},
    note = {R package version 2.3-40},
    url = {http://CRAN.R-project.org/package=raster},
  }

XX Refer in the man files more to the paper that I submit and state
that the occurrence sites and background sites are from that paper and
refere to Fucus distichus occurrence sites.

XX Jueterbock2015 Newly submitted manuscript

XXAkaike, 1974
XX Fielding and Bell, 1997
xx Warren et al., 2010
XX Warren and Seifert, 2011
XX Jimenez-Valverde, 2012
